---
title: "California Housing Prices"
author: "Robert Castillo"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#EDA
```{r}
library(car)
library(randomForest)
```


```{r}
data <- read.csv('California_Houses.csv')
```

```{r}
head(data)
```

```{r}
summary(data)
```


```{r}
# Median House Value
qqnorm(data$Median_House_Value, pch =1, frame=FALSE)
qqline(data$Median_House_Value, col = 'steelblue', lwd=2)
```

```{r}
symbox(~Median_House_Value, data= data)
```
```{r}
hist(sqrt(data$Median_House_Value))
qqnorm(sqrt(data$Median_House_Value), pch =1, frame=FALSE)
qqline(sqrt(data$Median_House_Value), col = 'steelblue', lwd=2)
```

```{r}
#Median Income
qqnorm(data$Median_Income, pch =1, frame=FALSE)
qqline(data$Median_Income, col = 'steelblue', lwd=2)
```

```{r}
symbox(~Median_Income, data= data)
```

```{r}
hist(log(data$Median_Income))
qqnorm(log(data$Median_Income), pch =1, frame=FALSE)
qqline(log(data$Median_Income), col = 'steelblue', lwd=2)
```

```{r}
# Median Age
qqnorm(data$Median_Age, pch =1, frame=FALSE)
qqline(data$Median_Age, col = 'steelblue', lwd=2)
```

```{r}
symbox(~Median_Age, data= data)
```

```{r}
#Total Rooms
qqnorm(data$Tot_Rooms, pch =1, frame=FALSE)
qqline(data$Tot_Rooms, col = 'steelblue', lwd=2)
```

```{r}
symbox(~Tot_Rooms, data= data)
```

```{r}
hist(log(data$Tot_Rooms))
qqnorm(log(data$Tot_Rooms), pch =1, frame=FALSE)
qqline(log(data$Tot_Rooms), col = 'steelblue', lwd=2)
```

```{r}
#Tot_Bedrooms
qqnorm(data$Tot_Bedrooms, pch =1, frame=FALSE)
qqline(data$Tot_Bedrooms, col = 'steelblue', lwd=2)
```

```{r}
symbox(~Tot_Bedrooms, data= data)
```

```{r}
hist(log(data$Tot_Bedrooms))
qqnorm(log(data$Tot_Bedrooms), pch =1, frame=FALSE)
qqline(log(data$Tot_Bedrooms), col = 'steelblue', lwd=2)
```

```{r}
#Population
qqnorm(data$Population, pch =1, frame=FALSE)
qqline(data$Population, col = 'steelblue', lwd=2)
```

```{r}
symbox(~Population, data= data)
```

```{r}
hist(log(data$Population))
qqnorm(log(data$Population), pch =1, frame=FALSE)
qqline(log(data$Population), col = 'steelblue', lwd=2)
```

```{r}
#Households
qqnorm(data$Households, pch =1, frame=FALSE)
qqline(data$Households, col = 'steelblue', lwd=2)
```

```{r}
symbox(~Households, data=data)
```

```{r}
hist(log(data$Households))
qqnorm(log(data$Households), pch =1, frame=FALSE)
qqline(log(data$Households), col = 'steelblue', lwd=2)
```

```{r}
# Distance to Coast
qqnorm(data$Distance_to_coast, pch =1, frame=FALSE)
qqline(data$Distance_to_coast, col = 'steelblue', lwd=2)
```

```{r}
symbox(~Distance_to_coast, data = data)
```

```{r}
hist(log(data$Distance_to_coast))
qqnorm(log(data$Distance_to_coast), pch =1, frame=FALSE)
qqline(log(data$Distance_to_coast), col = 'steelblue', lwd=2)
```

```{r}
city_dists <- data[c('Distance_to_LA', 'Distance_to_SanDiego', 'Distance_to_SanJose', 'Distance_to_SanFrancisco')]

closest <- apply(city_dists, 1, FUN = which.min)

f_closest <- as.factor(closest)
levels(f_closest) <- c('LA', 'SD', 'SJ', 'SF')

barplot(table(f_closest), main='Freq of Closest Major Cities')
```


```{r}
data2 <- data

data2$Median_House_Value_Sqrt <- sqrt(data2$Median_House_Value)
data2$Median_Income_Log <- log(data2$Median_Income)
data2$Distance_to_coast_Log <- log(data2$Distance_to_coast)

data2$Closest_City <- f_closest
```

```{r}
data2 <- data2[-c(1,2,8:14)]
```


**Residual Plots**

```{r}
#Median_Age
lm_age <- lm(Median_House_Value_Sqrt~ Median_Age, data = data2)
plot(lm_age)
```

```{r}
#Tot_Rooms
lm_rooms <- lm(Median_House_Value_Sqrt ~ Tot_Rooms, data = data2)
plot(lm_rooms)
```

```{r}
#Tot_Bedrooms
lm_bedrooms <- lm(Median_House_Value_Sqrt ~ Tot_Bedrooms, data = data2)
plot(lm_bedrooms)
```

```{r}
#Population
lm_pop <- lm(Median_House_Value_Sqrt ~ Population, data = data2)
plot(lm_pop)
```

```{r}
#Households
lm_house <- lm(Median_House_Value_Sqrt ~ Households, data = data2)
plot(lm_house)
```

```{r}
#Median_Income_Log
lm_income <- lm(Median_House_Value_Sqrt ~ Median_Income_Log, data = data2)
plot(lm_income)
```

```{r}
#Distance to coast Log
lm_distance <- lm(Median_House_Value_Sqrt ~ Distance_to_coast_Log, data=data2)
plot(lm_distance)
```

```{r}
n <- nrow(data2)
train_index <- sample(n, round(n*.8))
test_index <- (1:n)[-train_index]
rf_model <- randomForest(Median_House_Value_Sqrt ~ ., data = data2[train_index, ], mtry = 4, importance = TRUE)
print(rf_model)
```

```{r}
varImpPlot(rf_model)
```

```{r}
m1 <- lm(Median_House_Value_Sqrt ~ Median_Income_Log + Distance_to_coast_Log + Median_Age + Population + Closest_City, data = data2)
summary(m1)
```

```{r}
vif(m1)
```

#Modeling
```{r}
housing_df <- read.csv('California_Houses.csv')

# eliminate 500001 values due to the fact that the data collection process seemed to 
# have set all values greater than 500K equal to 500001 so we have a very large number of 
# observations with 500001 as the price (even if the other features change)
housing_df <- housing_df[housing_df$Median_House_Value < 500001, ]
```
&nbsp;&nbsp;

Summary of response variable (Median_House_Value) and the predictors is shown above. 


## Add to engineered features

&nbsp;&nbsp;
```{r}

# closest major city
city_dists <- housing_df[c('Distance_to_LA',	'Distance_to_SanDiego',	'Distance_to_SanJose',	'Distance_to_SanFrancisco')]
closest <- apply(city_dists, 1, FUN = which.min)
f_closest <- as.factor(closest)
levels(f_closest) <- c('LA',	'SD',	'SJ',	'SF')


# dist to coast
dist2coast <- housing_df['Distance_to_coast']
dist2coast <- sapply(dist2coast, as.numeric)
coast_levels <- cut(dist2coast,c(0, 3200, 9079.8, 20522.0, 150000, 333804.7))
levels(coast_levels) <- c("On", "Very Near", "Near", "Far", "Very Far" )

# age (simpler cut proved a bit more significant)
medAge <- housing_df['Median_Age']
medAge <- sapply(medAge, as.numeric)
ages <- cut(medAge,c(0, 45, max(medAge)))
levels(ages) <- c("Not Old", "Old")

# add interaction features (simpler cuts from EDA)

avg_rooms_per_house <- housing_df$Tot_Rooms / housing_df$Households
avg_rooms <- cut(avg_rooms_per_house,c(0, 6, max(avg_rooms_per_house)))
levels(avg_rooms) <- c("Less than 6", "6 or more" )
avg_rooms_per_house[avg_rooms_per_house  >= 100] = NA


avg_people_per_house <- housing_df$Population / housing_df$Households
avg_people <- cut(avg_people_per_house,c(0, 2, 3, max(avg_people_per_house)))
levels(avg_people) <- c("Less Than 2", "2 to 3", "More than 3" )
avg_people_per_house[avg_people_per_house  >= 100] = NA


avg_people_per_room <- housing_df$Population / housing_df$Tot_Bedrooms
avg_ppr <- cut(avg_people_per_room,c(0, 1, 3, max(avg_people_per_room)))
levels(avg_ppr) <- c("Less Than 1", "1 to 3", "More than 3" )
avg_people_per_room[avg_people_per_room  >= 10] = NA

# factors added to df 
housing_df$closest_city <- f_closest
housing_df$coast_levels <- coast_levels
housing_df$ages_levels <- ages
housing_df$rooms_per_house_levels <- avg_rooms
housing_df$people_per_house_levels <- avg_people
housing_df$people_per_room_levels <- avg_ppr

# numeric interactions added to df
housing_df$avg_rooms_per_house <- avg_rooms_per_house  
housing_df$avg_people_per_house <- avg_people_per_house  
housing_df$avg_people_per_room <- avg_people_per_room  

# apply transformation for new (approx. normal) response variable
housing_df$Median_House_Value_Sqrt <- sqrt(housing_df$Median_House_Value)

```
&nbsp;&nbsp;


# Modeling without Interactions

## Random Forest Model for Baselines

&nbsp;&nbsp;

```{r, eval=FALSE}

# with final cuts of categorical variables
rf_model <-randomForest(Median_House_Value_Sqrt~ . - Median_House_Value - Latitude
                   - Longitude - Median_Age - Distance_to_SanDiego 
                   - Distance_to_SanJose - Distance_to_SanFrancisco
                   - Distance_to_coast - Distance_to_LA, 
                   data=na.omit(housing_df),
                   ntree=500) 
print(rf_model)

```

&nbsp;&nbsp;

# Get Estimate for Test Error Using 10 fold CV 


&nbsp;&nbsp;

```{r, eval=FALSE}

set.seed(7)

train.control <- trainControl(method = "cv", number = 10)

repGrid <- expand.grid(.mtry=c(8)) # standard default parameter


cv_rf <- train(Median_House_Value_Sqrt ~ Median_Income + Tot_Rooms + 
                 Households + closest_city + coast_levels + ages_levels + 
                 rooms_per_house_levels + people_per_house_levels + people_per_room_levels + 
                 avg_rooms_per_house + avg_people_per_house + avg_people_per_room, 
                 data = na.omit(housing_df), 
                 method = "rf", 
                 trControl = train.control,
                 ntree=300, 
                 tuneGrid = repGrid)
  
print(cv_rf)


#Random Forest 

#19643 samples
#   12 predictor

#No pre-processing
#Resampling: Cross-Validated (10 fold) 
#Summary of sample sizes: 17678, 17680, 17679, 17680, 17678, 17678, ... 
#Resampling results:

#  RMSE      Rsquared   MAE     
#  61.21744  0.6893869  45.43543

#Tuning parameter 'mtry' was held constant at a value of 8

```

&nbsp;&nbsp;


## MLR Modeling 

We use backward selection on no interaction model:

&nbsp;&nbsp;

```{r}
model_no_int <- lm(Median_House_Value_Sqrt ~ . - Median_House_Value - Latitude
                   - Longitude - Median_Age - Distance_to_SanDiego 
                   - Distance_to_SanJose - Distance_to_SanFrancisco
                   - Distance_to_coast - Distance_to_LA,
                   data = na.omit(housing_df))
summary(model_no_int) 

# use backward selection
model_no_int2 <- step(model_no_int, direction = "backward", trace=0) 
summary(model_no_int2)



# no significant difference in model with extra terms
anova(model_no_int2, model_no_int)


# get rid of seemingly unnecessary variables 
model_no_int3 <- lm(formula = Median_House_Value_Sqrt ~ Median_Income + Tot_Rooms
                    + Households + closest_city + coast_levels + ages_levels
                    + rooms_per_house_levels + people_per_house_levels + people_per_room_levels, 
                    data = na.omit(housing_df))
summary(model_no_int3)


# significant difference in model with extra terms, so use model_no_int2
anova(model_no_int3, model_no_int2)
```









